---
title: "data_import"
author: "AKW"
date: "2025-03-24"
output: html_document
---


```{r}
library(readxl)
file_path<-"C:/Users/USER/Documents/repository/data/Outcome.csv"
DATA_DUMMY <- read.csv(file_path)
summary(DATA_DUMMY)

```
<<<<<<< HEAD


=======
>>>>>>> origin/main
```{r}
# Convert all character columns in DATA_DUMMY to factors to ease summary statistics

DATA_DUMMY[] <- lapply(DATA_DUMMY, function(x) {
  if(is.character(x)) as.factor(x) else x
})
library(dplyr)
library(knitr)
# Summarize the number of new cancer cases by year and cancer type
summary_by_year_and_cancer_type <- DATA_DUMMY %>%
  group_by(YearofDiagnosis, TypeofGyneCancer) %>%
  summarise(total_cases = n()) %>%
  arrange(YearofDiagnosis, TypeofGyneCancer)

# View the summarized data
kable(summary_by_year_and_cancer_type)
```
<<<<<<< HEAD



```{r}

# Calculate deaths annually and by cancer type
deaths_by_year_and_cancer_type <- DATA_DUMMY %>%
  filter(Outcome == "Dead") %>%  # Filter for deaths
  group_by(YearofDiagnosis, TypeofGyneCancer) %>%
  summarise(total_deaths = n()) %>%   # Count deaths per year and cancer type
  arrange(YearofDiagnosis, TypeofGyneCancer)

# View the summarized data
kable(deaths_by_year_and_cancer_type)


```

#Tabulating the incidence by cancer type
```{r}
# Separate tables for each type of cancer
# Load tidyr package
library(tidyr)

# Combine into one table with separate columns for each cancer type
combined_cancer_cases <- summary_by_year_and_cancer_type %>%
  pivot_wider(names_from = TypeofGyneCancer, 
              values_from = total_cases, 
              values_fill = 0)  # Fill with 0 for missing values

# View the combined table
kable(combined_cancer_cases)

```



#plotting the trend of the cancers from 2018 to 2024
```{r}
# Load ggplot2 package
library(ggplot2)

# Create the line graph
ggplot(summary_by_year_and_cancer_type, aes(x = YearofDiagnosis, 
                                            y = total_cases, 
                                            color = TypeofGyneCancer, 
                                            group = TypeofGyneCancer)) +
  geom_line(size = 1) +  # Create the line graph
  geom_point(size = 2) +  # Add points for better visibility
  labs(title = "Incidence of Gynecological Cancer Cases by Year and Cancer Type", 
       x = "Year of Diagnosis", 
       y = "Number of Cases", 
       color = "Cancer Type") +
  theme_minimal() +
  theme(legend.position = "right")


```

#cumulative cases vs deaths by gynecologic cancer type
```{r}
# Step 1: Summarize incidence and deaths by cancer type
# Incidence
incidence_summary <- DATA_DUMMY %>%
  group_by(TypeofGyneCancer) %>%
  summarise(total_incidence = n())

# Deaths
death_summary <- DATA_DUMMY %>%
  filter(Outcome == "Dead") %>%
  group_by(TypeofGyneCancer) %>%
  summarise(total_deaths = n())

# Step 2: Merge incidence and death data
cancer_summary <- full_join(incidence_summary, death_summary, by = "TypeofGyneCancer")



# Replace any NA values (cases where no deaths were recorded) with 0
cancer_summary[is.na(cancer_summary)] <- 0
print(cancer_summary)

# Step 3: Modify data for back-to-back plotting
# Make the incidence negative for plotting on the left side
cancer_summary <- cancer_summary %>%
  mutate(total_incidence = -total_incidence)

# Step 4: Plot the back-to-back bar chart with relabeled x-axis
ggplot(cancer_summary, aes(y = TypeofGyneCancer)) +
  geom_bar(aes(x = total_incidence), stat = "identity", fill = "blue") +  # Incidence (left side)
  geom_bar(aes(x = total_deaths), stat = "identity", fill = "red") +      # Deaths (right side)
  labs(title = "No. of cases                               Deaths",
       x = "Count",
       y = "Cancer Type") +
  theme_minimal() +
  coord_cartesian(xlim = c(-300, 300)) +
  scale_x_continuous(labels = function(x) abs(x),   # Relabel x-axis to show positive values
                     breaks = seq(-max(abs(cancer_summary$total_incidence)), max(cancer_summary$total_deaths), by = 50)) + 
  theme(plot.title = element_text(hjust = 0.5))




```


# Summarize cancer cases by different indicators
```{r}
# Ensure DATA_DUMMY is a data frame/tibble
DATA_DUMMY <- as_tibble(DATA_DUMMY)

# Group by education level, calculate frequency, and percentage
freq_by_education <- DATA_DUMMY %>%
  group_by(LevelofEduc) %>%
  summarise(frequency = n()) %>%
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Create a new column combining frequency and percentage
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

freq_by_education %>%
  select(LevelofEduc, frequency_with_percentage) %>%
  kable(col.names = c("Education Level", "Frequency (Percentage)"))


# Group by age group, calculate frequency, and percentage
freq_by_age <- DATA_DUMMY %>%
  mutate(age_group = case_when(
    Age < 20 ~ "<20",
    Age >= 20 & Age< 30 ~ "20-29",
    Age >= 30 & Age < 40 ~ "30-39",
    Age >= 40 & Age < 50 ~ "40-49",
    Age >= 50 & Age < 60 ~ "50-59",
    Age >= 60 & Age < 70 ~ "60-69",
    Age >= 70 ~ ">70"
  )) %>%
  group_by(age_group) %>%
  summarise(frequency = n()) %>%
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Combine frequency and percentage into a single column
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

# Display the table with kable, showing combined frequency and percentage
freq_by_age %>%
  select(age_group, frequency_with_percentage) %>%
  kable(col.names = c("Age Group", "Frequency (Percentage)"))



# Group by marital status and calculate frequency
freq_by_marital <- DATA_DUMMY %>%
  group_by(Marital_status) %>%
  summarise(frequency = n()) %>%
  # Calculate percentage
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Create a new column combining frequency and percentage
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

# Display the table with kable, showing frequency with percentage in brackets
freq_by_marital %>%
  select(Marital_status, frequency_with_percentage) %>%
  kable(col.names = c("Marital Status", "Frequency (Percentage)"))



# Group by Subcounty, calculate frequency, and percentage
freq_by_location <- DATA_DUMMY %>%
  group_by(Subcounty) %>%
  summarise(frequency = n(), .groups = 'drop') %>%
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Combine frequency and percentage into a single column
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

# Display the table with kable, showing combined frequency and percentage
freq_by_location %>%
  select(Subcounty, frequency_with_percentage) %>%
  kable(col.names = c("Sub County", "Frequency (Percentage)"))


# Group by Treatment type, calculate frequency, and percentage
freq_by_treatment <- DATA_DUMMY %>%
  group_by(Treatment) %>%
  summarise(frequency = n(), .groups = 'drop') %>%
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Combine frequency and percentage into a single column
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

# Display the table with kable, showing combined frequency and percentage
freq_by_treatment %>%
  select(Treatment, frequency_with_percentage) %>%
  kable(col.names = c("Treatment Type", "Frequency (Percentage)"))


```


# Summarize the number of cases by cancer stage
```{r}
summary_by_stage <- DATA_DUMMY %>%
  group_by(Clinical_stage) %>%
  summarise(total_cases = n()) %>%
  arrange(desc(total_cases))

# View the summarized data
kable(summary_by_stage)


# Summarize the number of cases by cancer stage and type
summary_by_stage_and_type <- DATA_DUMMY %>%
  group_by(Clinical_stage,TypeofGyneCancer) %>%
  summarise(total_cases = n()) %>%
  ungroup()  # Ungroup to avoid grouping issues

# Reshape the data into a wide format for better readability
summary_stage_wide <- summary_by_stage_and_type %>%
  pivot_wider(names_from = TypeofGyneCancer, 
              values_from = total_cases, 
              values_fill = 0)  # Fill missing values with 0 for empty combinations

# View the reshaped table
kable(summary_stage_wide)


```

```{r}

# Group and summarize the data
cancer_by_education <- DATA_DUMMY %>%
  group_by(LevelofEduc, TypeofGyneCancer) %>%
  summarise(total_cases = n(), .groups = 'drop') %>%
  pivot_wider(names_from = TypeofGyneCancer, 
              values_from = total_cases, 
              values_fill = 0) %>%
  # Calculate the total number of cases for each education level
  mutate(Total = rowSums(select(., where(is.numeric)))) %>%
  # Calculate the percentage for each cancer type
  mutate(across(where(is.numeric), 
                ~ paste0(.x, " (", round(.x / Total * 100, 1), "%)"))) 

# Print clean table with counts and percentages
cancer_by_education %>%
  kable()
```



```{r}


# Group by marital status and calculate frequency
freq_by_marital <- DATA_DUMMY %>%
  group_by(Marital_status) %>%
  summarise(frequency = n()) %>%
  # Calculate percentage
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Create a new column combining frequency and percentage
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

# Display the table with kable, showing frequency with percentage in brackets
freq_by_marital %>%
  select(Marital_status, frequency_with_percentage) %>%
  kable(col.names = c("Marital Status", "Frequency (Percentage)"))


```

# Objective two
# part one;
# Summarize cancer cases by subcounty and cancer type
```{r}
cancer_by_subcounty <- DATA_DUMMY %>%
  group_by(Subcounty,TypeofGyneCancer) %>%
  summarise(total_cases = n()) %>%
  pivot_wider(names_from = TypeofGyneCancer, 
              values_from = total_cases, 
              values_fill = 0)  # Fill missing values with 0

# View the result
kable(cancer_by_subcounty)



```

## Objective Three
#  COMORBIDITIES
```{r}

# Group by Comorbidities, calculate frequency, and percentage
freq_by_location <- DATA_DUMMY %>%
  group_by(Comorbidities) %>%
  summarise(frequency = n(), .groups = 'drop') %>%
  mutate(percentage = round((frequency / sum(frequency)) * 100, 1)) %>%
  # Combine frequency and percentage into a single column
  mutate(frequency_with_percentage = paste0(frequency, " (", percentage, "%)"))

# Display the table with kable, showing combined frequency and percentage
freq_by_location %>%
  select(Comorbidities, frequency_with_percentage) %>%
  kable(col.names = c("Cormonidities", "Frequency (Percentage)"))



```
=======
>>>>>>> origin/main
